---
title: "homework4"
author: "Anders Hjulmand"
date: "9/21/2021"
output:
  pdf_document: default
  html_document: default
  word_document: default
---
### Github: https://github.com/Digital-Methods-HASS/au651039_hjulmand_anders

```{r}
#loading packages
pacman::p_load(tidyverse)
```


### 1) Use R to figure out how many elements in the vector below are greater than 2.

```{r}
#Defining the vector
rooms <- c(1, 2, 1, 3, 1, NA, 3, 1, 3, 2, 1, NA, 1, 8, 3, 1, 4, NA, 1, 3, 1, 2, 1, 7, 1, NA)

#summing values above 2. Removing NA's.
sum(rooms > 2, na.rm=T)
```
### 2) What type of data is in the 'rooms' vector?
The rooms-vector contain integers and missing values called NA's. The type of vector is a double vector 
```{r}
typeof(rooms)
```

### 3) What is the result of running the median() function on the above 'rooms' vector?
```{r}
#trying to run the median() on rooms
median(rooms)
#It gives the output NA

#running median() on rooms and excluding NA's
median(rooms, na.rm=T)

#The median of rooms is 1.5
```

### 4) Submit the following image to Github: Inside your R Project (.Rproj), install the 'tidyverse' package and use the download.file() and read_csv() function to read the SAFI_clean.csv dataset into your R project as 'interviews' digital object (see instructions in https://datacarpentry.org/r-socialsci/setup.html and 'Starting with Data' section). Take a screenshot of your RStudio interface showing

a) the line of code you used to create the object,
```{r, message = F}
df_safi <- read_csv("SAFI_clean.csv")
```
b) the 'interviews' object in the Environment, and

c) the file structure of your R project in the bottom right "Files" pane.

Save the screenshot as an image and put it in your AUID_lastname_firstname repository inside our Github organisation (github.com/Digital-Methods-HASS) or equivalent. Place here the URL leading to the screenshot in your repository.
